XRONOS: an open data infrastructure for archaeological chronology

### 2024-07-29 paper_jcaa.qmd
Rewrite for JCAA

---
title: "XRONOS: an open infrastructure for archaeological chronometric data"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    email: martin.hinz@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    country: Switzerland
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    country: Germany
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
keywords: 
  - open data
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
  - chronological modelling
bibliography: references.bib
format:
  hikmah-pdf:
    papersize: a4
    mainfont: "Linux Libertine"
    sansfont: "URW Gothic"
    biblatex-chicago: true
    biblio-style: authordate
    biblatexoptions:
      - backend=biber
      - autolang=hyphen
      - isbn=false
      - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
---

<!-- 
For submission to Archaeometry. Guidelines:

- Max 8000 words including all figures etc.
- Max 6 figures/tables
- Abstract max 100 words
- Max 10 keywords
- Word 2002 or earlier or RTF (???)
- Tables seperately in XLS
- Figures seperately in TIFF or EPS
- double spaced, wide

OR as a JCAA 'case study' - 5,000 words
  -->

@@

* Chronometric modelling depends on ready access to chronometric data!
* Luckily, we do have a lot of it.
* Moving from 'open data' (uploading a CSV) to 'open infrastructure' [@Kintigh2006]

Chronometric data has been one of the principle domains in which open data and open science approaches have been applied over the last ten years. 
Comprehensive compilations of radiocarbon dates have become available for many parts of the world in the last decade and, as natural next step, there are now several initiatives to collate this data globally, including the retrieval tool c14bazAAR [@SchmidEtAl2019], the IntChron exchange format [@BronkRamseyEtAl2019], and the synthetic database p3kc14 [@BirdEtAl2022]. 
This move from archiving individual datasets to producing a living open data infrastructure for archaeological chronologies raises a number of challenges. 
Radiocarbon datasets are still sorely lacking for many parts of the world and, even in those regions with good coverage, the quality of data is highly uneven and largely undocumented. 
There exists no central repository ensuring the long-term sustainability and completeness of these datasets, and the potential of placing other sources of chronometric information (e.g. dendrochronology, typological dating) in an open data framework has hardly been realised at all.

Here, building on and complementing these initiatives, we present XRONOS: a new web-based platform for chronometric data from archaeological contexts worldwide, combining an open data repository with tools for importing, curating and analysing chronometric information from diverse sources. 
The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; 
develop robust tools for the continuous ingestion and refinement of this data; 
and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.

This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2024), including a generalised data model for site and radiocarbon information, extendable to other chronometric data; 
an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; 
a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
and an R package for interfacing with the API. 
We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.

In this paper we:

* Review the state of the art in radiocarbon data compilation (and other chronometric data, such as they are), highlighting the need for XRONOS
* Describe XRONOS' design goals and software architecture.
* Describe the implementation of some representative features
  * With a focus on the choices we made in representing/modelling chronological data
* Assess the current state of the software and database, and give case studies of how it can be used
* Conclude with our future plans

@@

We live in interesting times: Today, archaeological research can no longer complain that too few absolute dates exist for the phenomena of the past. Large-scale dating projects such as \[example\] and \[example\] are producing extensive data series in a short time, which make it possible to situate individual sites in time with unprecedented precision. But also due to the dating methods that have become affordable even for smaller budgets in the meantime and the integration of natural scientific dating that is almost mandatory, such large initiatives are accompanied by a great amount of individual and scattered chronological information that represents a rich source of data for supra-regional investigations. Based on that, data modelling can be used as a method to draw conclusions from a variety of individually information-weak data.

The problem, however, is that these data are very scattered. Traditionally, radiocarbon data were disseminated by the laboratories, with special attention paid to publication in the form of "date lists" in the journal *Radiocarbon*. Today, unfortunately, this has become very rare, on the one hand because the quantity and complexity of the data limit the possibilities of publication on paper. On the other hand, data in printed form are hardly suitable for reuse today. As a result, access to chronological information, especially a comprehensive overview, is made much more difficult [@BronkRamseyEtAl2019]. Furthermore, data is presented in a variety of different formats and granularities, and critical information is often missing, although standards exist [@Millard2014].

Without data repositories, most comprehensive analyses are no longer feasible. So far, there has been no overarching database that constantly collects both 14C and dendrochronological data and makes them accessible in the long term. In the spirit of Open Science, it is the goal of XRONOS to establish such an Open Access database. Our primary ambition is to collect 14C and dendrochronological data for Switzerland and Europe as extensively and comprehensively as possible. However, the project is designed in such a way that an expansion beyond the originally defined spatial focus is already a reality.

Currently, all scientific disciplines face a revolution in the way data is disseminated and used [@BronkRamseyEtAl2019].
The access to these data, and methods integrate individual data into larger scientific questions of social relevance are both are decisive, but not yet sufficiently implemented prerequisites to reach an understanding of the temporal phenomena and dynamics in past societies [@Kristiansen2014]. 
This novel, cross-disciplinary circulation and utilisation makes it possible to deal with completely new scientific questions and makes archaeological data relevant in thoroughly new contexts. 
For example, analysis of past climatic changes and their long-term effects rely on information from archaeologists to estimate the impact on human and biotic activities [@KintighEtAl2015, 7-8]<!-- From Antiquity paper, not sure which paper exactly this is -->.
An important source of such information is absolute chronological data.
However, the basis of any empirically based modelling is the availability of data.

@@

* Where do radiocarbon dates come from?
* Where are they published?

Although 'open data' is often thought to be a new phenomenon in archaeology [@CITE], it has a long history in radiometric dating.
Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical: the last 'date list' appeared in 19XX, and unfortunately was not replaced by another form of systematic data-sharing or dissemination.
Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
Thus, even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice the only way to produce aggregated datasets recently has been to manually search through relevant literature for dates reported by the submitter of the sample.
This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions [@Millard2014], the lack of any convention on the reporting of contextual information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.

* Graph of number of dates/databases through time
* Table of databases

Despite these difficulties, there have been many efforts to systematically compile radiocarbon dates in this way over the last few decades [@FIG].

E.g. RADON:
At the time, this was one of the largest 14C data collections for absolute chronological information in Europe. 
RADON and RADON-B can be considered a complete success, as they have been the basis for several of our own publications (Hinz, 2015; Hinz, Feeser, et al., 2012; Krossa et al., 2017; Weinelt et al., 2015) and as a data repository for high-level publications by other research groups (Bevan et al, 2017, 2019; Dolukhanov and Shrukov, 2004; Henkner et al., 2018; Lemmen and Wirtz, 2014; Manning et al, 2016; Pinhasi et al, 2005; Puchol and García, 2017; Sanders, 2012; Shennan et al, 2013; Timpson et al, 2014), and inspired other projects (Schmid, 2018; IsoMemo; Schmid et al, 2018).

We are aware of XX existing radiocarbon databases or published datasets, comprising an estimated XXXX individual records [@TABLE] – though this is likely an underestimate.
The majority of these resources have a specific regional (XX%) or temporal (XX%) scope.
Others are associated with individual radiocarbon labs (e.g. KIK-IRPA, ORAU, BANADORA) and a handful aim at global coverage (XX%, discussed further below).

* How are they published? As databases, SI, etc.?
* How many are open?

While these compilations have greatly advanced access to chronometric data and enabled new modes of analysis (e.g. 'dates as data', @CITE), there are some obvious limitations to the fragmentary way in which they have been compiled.

* And patchy: highly uneven spatial and geographic coverage, different data collected, different conventions for recording data
* Regional databases are put together *ad hoc* and rarely continuously maintained over long periods
* The coverage of these databases is uneven, systematically biased, and duplicative of each other
* Not all are open
* Not all are machine readable

* Recent moves towards a global synthesis: c14bazAAR [@SchmidEtAl2019], IntChron [@BronkRamseyEtAl2019], p3k14c [@BirdEtAl2022]
  * Each taking a different approach, pros and cons
  * We see a need for something that is comprehensive, unified, living
* And we don't see why it should be limited to radiocarbon: what about other radiometric dates? Dendro? Typologies?

* Some databases, such as KIK-IRPA, ORAU and BANADORA, which are maintained by 14C laboratories, only contain their own measurement data and are therefore of very limited use for overarching research projects.
* Others, such as CONTEXT, RADON, RADON-B, 14SEA and CARD, have a limited regional or temporal focus, and the rich data available e.g. in Switzerland and surrounding countries is not sufficiently represented in these databases. In addition, some of them are project-specific and their continued existence can be described as uncertain due to funding constraints.
* Recently, [@BronkRamseyEtAl2019] introduced IntChron, a data structure designed to ensure easy availability of radiocarbon data from published articles. We think this approach is very commendable and it inspires us in terms of reproducibility of dating. But we want to go beyond that and put more emphasis on the evaluation and dissemination of relevant information. "IntChron is not intended to be a stand-alone database" [@BronkRamseyEtAl2019, 5], but XRONOS is. An overarching database is critical for any research application in terms of data retrievability. Nevertheless, interoperability with other data services, including IntChron in particular, is very important to us.

@@

* There are lots of other types of chronometric data!
  * Dendro - in certain places (e.g. Switzerland!), way more important than radiocarbon
  * Other radiometric/absolute dating methods, OSL, etc.
  * Typology - still the backbone of many chronologies
  * Application-specific: sea level dating, rock patina, etc.
* Compilation of other types of dates has been much more limited.
  * Exceptions: ...?

<!-- MH to do: -->
For European dendrochronological data, the situation is even less favourable for archaeological purposes:

* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
* DendroDB has a European focus but only provides data for historical periods.
* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD). An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g. ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%). Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).

* Typology
  * perio.do
  * Often embedded in other types of database (e.g. gazetteers)
  * But nobody has treated it as a form of chronometric data in its own right?

What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data. However, to efficiently conduct data driven modelling of human-environmental dependencies and dynamics on the past, such a data base seems indispensable for future research.

@@

XRONOS builds on the structure of RADON (@RADON), one of the earliest comprehensive databases of radiocarbon dates.

@@

Our concrete goals in the design and implementation of XRONOS are the following:

1. Provision of access to 14C and dendrochronological data with meta information on archaeological contexts via web-based Open Access frontend
2. Web frontend with processing functions (filters and queries, export to various data formats, 14C calibration & chronology formation) for in place analytical functions
3. machine readable interface --- API --- for direct access e.g. by open source statistical environments (eg. R) for analytical purposes or dendroanalytical software
4. licensing and versioning of individual data using eg. DOI
5. Storage in a repository with user and role-specific access rights as well as encryption for data privacy
6. Ensuring sustainable data security and usage through compliance with standards, appropriate backup solutions and high-availability technologies

Points 1-3 are now largely implemented and ready to be used. The results of such a search is easily exportable in formats that are simple and generally usable, and represent common, open standards (CSV, JSON). Such formats can be conveniently imported into common evaluation environments (R, Matlab, Python, OxCal, Chronomodel), or even better, directly accessed via the API. This can serve to establish a single gateway to the different sources of information and eliminate the necessity for data publishers to maintain information in multiple formats. With regard to the 14C data, this analytical functions include the live calibration of the individual dates, allowing the most recent calibration curve to be considered. In addition there are extended calibration possibilities like sum, as known from the standard software solutions (e.g. OxCal). Furthermore, the Bayesian calibration in relation to phase models (sequential calibration under consideration of stratigraphy) may be implemented as far as possible, whereby this will take place in close connection with the endeavours of the further development of corresponding R packages. For the dendrochronological data, especially the implementation of visualizations as well as the possibility to create mean curves and to identify impact phases will represent extended possibilities of immediate analysis. This portfolio is to be expanded in consultation with the user group, especially the state (cantonal) dendrochronological laboratories of Switzerland. More elaborated functionality should be made available primarily via an R Package, since this would be too complex for the user to be channelled via an web interface.

An increasing number of research funders legitimately demand that the data that is used in the course of the research they fund become available to other researchers. This disclosure enables new knowledge to be extracted through comparative studies, data mining, etc.; it also enables better transferability and reproducibility of the research process, which increases the quality of research. All data in XRONOS should be uniquely referenceable. Common publication practices, which so far led to ambiguities regarding the data, must be overcome. Therefore, each individual data set should be permanently archived in its current version, and changes resulting from updated information or changed boundary conditions (updated calibration curves) should be provided in a new version with versioned unique identification. In this way, it can be ensured that the exact state of a reference to a date can be unambiguously stored. Only in this way can the research based on these data truly become traceable. One means for this is a Digital object identifier (DOI), which combines a unique and machine-readable as well as persistent referencing with easy retrievability. This functionality is still to be implemented. In order to make such a sustainable data referencing possible, XRONOS strives for a cooperation with longterm data repositories, so that a permanent storage of the relevant information can be guaranteed. This ensures that the collected information will be available to the research and public community beyond the lifetime of XRONOS and thus beyond a possible end of the project.

The ultimate goal of XRONOS is to create an open data structure for the exchange and evaluation of absolute chronological data by the entire research community. However, the data management with regard to dendrochronological data in particular is associated with different interests of the institutions involved, so that compromises have to be made here. For all dendrochronological laboratories control over data disclosure is indispensable. These are mainly publication-related reasons and data is used as "exchange capital" between laboratories. However, in order to provide an incentive for central data management, it must be ensured that users retain full sovereignty over their data. To this end, this project plans to keep the data fully encrypted in such a way that even the database administrators have no insight into the data. For this purpose, methods of asymmetric cryptography (public/private key encryption) based on passwords will be integrated into the database, which make it possible to make data available to individual other users or to the general public.

We hope that the combination of easy and open data access, the provision of analytical functions within the web interface and by means of R packages, the sustainable, transparent and unambiguous data management, the long-term archiving already considered in the initial phase, together with the possibility for individual users to have full sovereignty over their data, will distinguish XRONOS as an outstanding data source and repository, and that we will thereby contribute to the sustainable provision and use of big data in archaeology and advance the use and improvement of standards in this field.

* Inspired by GBIF, which manages to integrate large amounts of data from diverse sources into a signal, consistent database

@@

TODO: Figure: conceptual scheme, centred on chrons (i.e. an idealised form of the database).

@@

* Bibliographic references
* Paper trail

* LOD??

### Software architecture

The XRONOS data model is implemented as a relational database using the free and open source database management system PostgreSQL.
However, apart from backups and other routine maintenance procedures, all interaction with the database is via a web application, which thus forms the core of XRONOS' architecture.
The XRONOS web application is written in Ruby and uses CRUD (Create, Read, Update, Delete) and MVC (Model, View, Controller) patterns as implemented in the Ruby on Rails framework.
The choice of this somewhat traditional or 'boring' architecture—as opposed to the non-relational/semantic technologies that are more *à la mode* in the digital humanities [@Fan2018; @Hyvonen2020; @SchloenProsser2023]—was purposeful, motivated by a sense that it offers a better chance of longevity and maintanability.
Our aim is for the software architecture to be as boring as possible so that the scientific contents can be as interesting as possible.

The XRONOS web application exposes two distinct user interfaces: a graphical user interface accessed through a web browser; and an application programming interface (API). 
Both interfaces follow a REST (Representational State Transfer) pattern [@VerborghEtAl2015], where each resource (e.g. a single radiocarbon date, a single bibliographic reference, or a single user) is statelessly mapped to a single address.
Users can then interact with resources at these addresses using a preditable and uniform interface based on HTTP verbs.
For example, the radiocarbon date RTD-8904 is represented by the address <https://xronos.ch/c14s/156205>.
Users can view information on this resource by sending a GET request to that address, regardless of which interface they are using, and authorised users can modify it using POST, PATCH, or DELETE requests.
The bibliographic reference associated with this date [@RichterEtAl2017] is similarly represented at the address <https://xronos.ch/references/17778>, and can be accessed at that address using the same interface as the radiocarbon date.
These uniform REST interfaces are another example of a boring architectural choice that make it easier for us to enrich the scientific contents of XRONOS, by  adding new types of modular resource that represent new scientific entities.

This basic REST pattern is augmented by seven 'actions' (following the standard pattern in Rails application) that express different ways of interacting with a resource: index, show, destroy, new, create, edit, and update. 
The 'show' action represents interaction with a single resource, as described above. 
The 'index' action, which lists resources of a given type (e.g. <https://xronos.ch/c14s> for radiocarbon dates), is worth special mention because it is through this that the filtering logic at the core of XRONOS' two interfaces is implemented.
By passing a query as HTTP GET parameters to the index action of a resource, the list returned the user is modified to only include records that match that query.
For example, <https://xronos.ch/sites?site[country_code]=CH> (the part of the URL after the `?` character encodes the SQL WHERE clause `country = 'CH'` as a GET parameter) lists sites in Switzerland <!-- TODO: check this actually works -->.
More complex queries can be executed using nested parameters.
For example, <https://xronos.ch/c14s?c14[sample][material][name]=charcoal> (encoding that the `c14` table should be joined to the `material` table via `sample`, followed by the WHERE clause `material.name = 'charcoal'`) lists radiocarbon dates obtained from charcoal samples <!-- TODO: check that this actually works... -->.
Uniquely, index actions can also respond with the result in a tabular data format (i.e. `.csv`).

The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g. at https://xronos.ch), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
Each action on each resource is represented by a page, though not all of these are publicly accessible.
Pages representing REST resources directly are supplemented by a number of synthetic interfaces, for example the 'data browser' (<https://xronos.ch/data>), which facilitates more complex filtering, or the search interface (<https://xronos.ch/search>).
The GUI also includes several resources which are not part of XRONOS' scientific data model, for example documentation pages, user profiles and news articles; these are currently not exposed in the API.

Access to various 'backstage' interfaces for creating, editing, and deleting data, and monitoring data quality is managed using a user permissions system.
Currently only authorised users affiliated with the XRONOS project can access these, but in the future we intend to support open registration and expose editing interfaces to all authenticated users.
For this reason, there is no sharp division between a 'public' and 'private' areas – viewing/querying data and editing/curating data share the same architecture and interface patterns.
<!-- TODO: say something about private data, even though it's not implemented yet? -->

The XRONOS API uses the same addresses as the web-based GUI (with the exception of some of the synthetic interfaces mentioned above) but responds with machine-readable data in JSON format, rather than a HTML page.
This response can be triggered by appending `.json` to the address or by including a HTTP `content-format` header in the request.
Though users can make such requests manually and parse the data with one of several off-the-shelf tools, the primary intended uses of this interface is to provide access for 1) programmatic clients to XRONOS and 2) other web services.
The XRONOS R package (discussed in more detail below) is an example of the former use, which could be extended to other programming environments used for scientific computing, e.g. Python or Haskell.
<!-- TODO: example of a web service that might use the API? -->

An overarching principle of this software architecture is that all interaction with XRONOS' data store, and as much of the data processing and 'business logic' of responding to REST requests as possible, is directed through the same server-side routines.
First and foremost, this allows us to provide multiple interfaces (i.e. the GUI and API, perhaps more in the future) without duplicating these elements of our codebase.
It also improves accessibility for users accessing XRONOS through devices with limited processing capability or through text-only browsers.
More broadly, avoiding reliance on client-side processing, e.g. with Javascript or Web Assembly (WASM)---which would be the other option---allows us keep our client interfaces simple (in most cases plain, semantic HTML pages and self-contained stylesheets) and therefore, we hope, sustainable in the face of constantly-evolving client-side technologies and standards.
It does have the weakness that, in practical terms, XRONOS is difficult to run and relies on the continued existence of a maintained external server.
We have however tried to mitigate this by providing clearly-documented source code and regular data dumps so that, if our instance of XRONOS disappears, or if one simply does not want to use it, it is possible for others to host a XRONOS server of their own.
As the sustainability of scientific software and data infrastructures is a pressing problem (a point we will return to in the conclusion), in the future it may be desirable to support further decentralisation through, for example, a federated server-server model [@CITE].

@@

The first prototype of XRONOS was developed in 2019.
It has been under sustained development 2021–2024 with the first public release in July 2021.

<!-- TODO: update/rewrite status of development -->
The current database is modular and, in its current stage of development, contains all the structures necessary for storing 14C data. 
In addition, the integration of dendrochronological information into the structure is already prepared, and a security and data protection architecture has been prepared for the final stage of development. 
The software is available as open source on Github (<https://github.com/xronos-ch),> and an R package for easy access, data retrieval and evaluation is also already available. 
XRONOS currently holds 92,129 14C data from 22,225 sites. 
The primary focus is on Switzerland and Europe, but the database also contains information from Africa, South America, Asia and Australia. 
The collection is subject to continuous addition.

Due to space limitations, and because we hope and expect XRONOS to continuously evolve, we don't attempt a comprehensive description of its implementation here.
Instead we offer a selection of representative features which illustrate our design goals and key aspects of our approach.

>>>>

---
title: "XRONOS: an open data infrastructure for archaeological chronology"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    email: joseph.a.roe@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    country: Switzerland
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    country: Germany
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It aims provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world. Collating a number of recent regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 date and 75,000 site records. It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating. Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it significantly easier to filter data of interest and reducing the need for manual data cleaning by end users. In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
keywords: 
  - open data
  - chronology
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
bibliography: references.bib
format:
  hikmah-pdf:
    papersize: a4
    mainfont: "Linux Libertine"
    sansfont: "URW Gothic"
    biblatex-chicago: true
    biblio-style: authordate
    biblatexoptions:
      - backend=biber
      - autolang=hyphen
      - isbn=false
      - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
---

<!--
* [ ] First draft:
  * [ ] Introduction [500]
  * [ ] State of the art [2000]
    * [ ] Compilations of radiocarbon dates [1000]
    * [x] Open data infrastructures [500]
    * [ ] Beyond radiocarbon [500]
  * [ ] Concept [1000]
    * [x] Inherits a common structure from RADON
    * [ ] Design goals [250]
    * [x] Data model [1150/500]
    * [x] LOD [250/250]
  * [x] Implementation [1000]
    * [x] Status of development [100/100]
    * [x] Software architecture [550/300]
    * [x] Data ingestion and integration [500/300]
    * [x] User interfaces [650/300]
  * [ ] Evaluation [500]
* [ ] Add references:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
* [ ] Figures and tables
* [ ] Copyedit:
  * [ ] Introduction
  * [ ] State of the art
  * [ ] Concept
  * [ ] Implementation
  * [ ] Evaluation
  -->

@@

* Chronology is the backbone of the archaeological record – we need it to be open!

<!-- Prehistory of open radiocarbon 
TODO: shorten -->
Although 'open data' is often thought to be a new phenomenon in archaeology [@CITE], it has a long history in radiometric dating.
Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical: the last 'date list' appeared in 19XX, and unfortunately was not replaced by another form of systematic data-sharing or dissemination.
Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
Thus, even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice the only way to produce aggregated datasets recently has been to manually search through relevant literature for dates reported by the submitter of the sample.
This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions [@Millard2014], the lack of any convention on the reporting of contextual information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.

* Move to open compilations in the last ~10 years
* Range of other chronological information used by archaeologists is largely untouched
* From uploading CSVs to (social) infrastructure

@@

@@

* Where do radiocarbon dates come from and how are they published?
* History of systematic compilations
* Summary of existing databases
  * FIGURE: graph of number of dates/databases through time
  * FIGURE: map of radiocarbon database coverage
  * TABLE: radiocarbon databases
* Problems with current state of the art

### Open data infrastructures

The profusion of radiocarbon compilations over the last decade has naturally prompted many to think more globally:

- The first available radiocarbon 'metadatabase' was **c14bazAAR** [@CITE], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (see @FIG).

- Another indexical approach is the **IntChron** project [@CITE], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common itnerface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
The JSON format also lends itself to the implementation of wrapper libraries, for example the 'rIntChron' package gives direct access to IntChron-indexed databases in R [@CITE].

* **p3k14c** [@CITE] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR. The major advantage of this approach is that the data is made internally consistent and has been manually cleaned to an extent, which makes it particularly well-suited to global analyses. The downside is that without the continuous link to the source databases present in the c14bazAAR and IntChron, it can only be kept up to date manually with periodic re-releases. An accompanying package (@CITE) provides direct access to the p3k14c dataset in R.

XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
But like p3k14c, it integrates these into a single database and applies data curation processes to harmonise them and improve the quality of the information.
It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
It also goes beyond the functionality of p3k14c by providing systems for the contunious ingestion and curation of new data as it is published.
However in general we see the approaches as complementary.
A c14bazAAR parser for XRONOS is in development (https://github.com/ropensci/c14bazAAR/pull/150), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
New data and corrections from p3k14c are incorporated into XRONOS as they are released.

XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
It parallels and draws inspiration from several similar initiatives in archaeology, such as <!-- TODO examples -->
And outside of it: <!-- TODO examples GBIF -->

@@

* Dendro
* Typology
* Other radiometric methods
* Application-specific 
* XRONOS aims to bring all these together, starting with radiocarbon, dendro-, and typology

@@

XRONOS inherits its basic structure from RADON (@CITE), with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for XX years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.

@@

<!-- TODO: Look at Martin's grant application and drafts -->

@@

* FIGURE: conceptual scheme

<!-- TODO: shorten -->

@@

Metadata is incorporated into XRONOS' data model at the level of the individual records (e.g. all records store their data of creation and last modification) and through two additional nodes: bibliographic references and versions.
Bibliographic references store information on the source of a record in the standard BibTeX (@CITE) format and can be linked through many-to-many associations to
sites or chrons.
Versions are a special type of record that are associated with all other records (including bibliographic references) and store the previous versions of those records as a series of changesets.
In this way all changes to data are recorded and can be reconstructed (or reversed) precisely.
This 'paper trail' also stores contextual metadata, e.g. who made the change and why.
It also means that records that are deleted can be reviewed or restored from their stored version history, which is never discarded.
Together these two systems provide a transparent record where the data in XRONOS comes from and how it has been altered, which we view as essential in a scientific data infrastructure.

### Linked data

The XRONOS data model presents several opportunities to link to other resources as linked open data (LOD) [@CITE].
We use controlled vocabularies in the form of the GBIF Backbone Taxonomy (for taxonomic descriptions of samples) via its API, and in future plan to extend this to for example site types (using the Getty and/or DAI<!-- TODO: check & cite -->).
Standardising field using these vocabularies to automatically then link the two entities.
Beyond this, there is also a concordance between XRONOS entities and other specialised data infrastructures in archaeology, such as Perio.do (@CITE, mapping to typological chrons) and gazetters like Pleiades or Vici (@CITE, mapping to site names).
Moving beyond archaeology, Wikidata (@CITE) already includes many of the concepts represented in the XRONOS model (e.g. archaeological sites, Q<!-- TODO -->).
Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the narrow field of archaeology.
It also allows us to enrich the database with contextual information that otherwise be beyond our scope and resources, for example embedding multilingual descriptions of sites from Wikipedia articles on site records where the site has been linked to a Wikidata item.

Conversely, we encourage others to use XRONOS as a linked open data resource by providing stable URLs and a machine-readable interface for every record.
Such usages could look like, for example, using a XRONOS URL as a canonical representation of a single radiocarbon date (e.g. <!-- TODO -->).
We also plan to implement an IntChron ([@CITE]) interface to XRONOS, to allow it to be indexed through that existing standard.<!-- TODO: also mention RDF? -->

@@

Following a short pilot project in 2019, the first phase of development of XRONOS was completed in 2021–2024, supported by a project grant from the Swiss National Science Foundation.
The web interface (<https://xronos.ch>) has been publicly accessible since July 2021.
Though we envisage XRONOS as a continuously-developing open source project and 'living database', the following offers a snapshot of progress at the end of our first grant-funded implementation phase.
We do not aim to be comprehensive, but rather to describe some key elements of XRONOS' current implementation that illustrate how our concept has been realised in practice.

### Software architecture
<!-- TODO: shorten -->
